Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-02 ms.tex
Removed commented text
clarification added on how to compute B_min

Galaxy clusters have for many years been proposed as sources of gamma rays. If shock acceleration in the ICM is an efficient process, a population of highly relativistic cosmic-ray (CR) hadrons, in the form of protons and heavy ions, in the ICM is to be expected. The main energy-loss mechanism for CR hadrons at high energies is pion production through the interaction of CRs with nuclei in the ICM. Pions are short lived and decay; the charged pions to electrons and positrons,\footnote{A charged pion decays to a muon or antimuon, which subsequently decay to an electron or positron as well as neutrinos.} and the neutral pion to gamma rays. Due to the low density of the ICM ($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large scales of a cluster and the volume filling magnetic fields in the ICM, CR hadrons will be confined in the cluster on timescales comparable and longer than the Hubble time \citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them to accumulate over time.  Depending on the CR pressure, quantified by the CR-to-thermal pressure fraction $X_\CR=\expval{P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma rays can be expected.

@@

A third mechanism for gamma-ray production in a galaxy cluster is self-annihilation of a DM particle, e.g. a weakly interacting massive particle (WIMP). As already mentioned, about 80\% of the cluster mass is in the form of dark matter. Recent works \citep{article:EvansFerrerSarkar:2004, article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM annihilations is expected to be proportional to the square of the DM density integrated along the line of sight, making clusters of galaxies potentially good
candidates for DM searches \citep{article:PinzkePfrommerBergstrom2009, 2011ApJ...726L...6C}. This despite their large distances compared to other common targets for DM searches, such as dwarf spheroidal galaxies \citep{article:Strigari_etal:2007, article:Acciari_etal:2010,article:Aliu_etal:2009} or the Galactic center \citep{article:Kosack_etal:2004, article:Aharonian_etal:2006, article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is one of the closest massive clusters \citep[$M \sim10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both a giant radio halo
\citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral radio relic, which appears connected to the radio halo with a ``diffuse'' bridge \citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested that the relic is an infall shock \citep{article:Ensslin_etal:1998} and successively demonstrated by cosmological simulations that model the non-thermal emission processes \citep{article:PfrommerEnsslinSpringel:2008, article:Pfrommer:2008, article:Battaglia_etal:2009, article:Skillman_etal:2011}. Extended soft thermal X-ray (SXR) emission is evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band \citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton \citep{article:Briel_etal:2001} revealed substructure in the X-ray halo with substantial turbulent pressure support that shows a lower limit of $\sim 10 \%$ of the total intracluster medium pressure \citep{2004A&A...426..387S}. As such it is a natural candidate for gamma-ray observations.

In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, with complementing analysis of available data from the Fermi Gamma-ray Space Telescope. The VERITAS and Fermi data have been used to place constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analyses, a present day Hubble constant of $H_{0} = 100h$ km s$^{-1}$ Mpc$^{-1}$ with $h=0.7$ has been used.

@@

\item In the last step, we scale the secondary radio emission profile linearly with $X_{\CR,0}$ and solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution would be degenerate.
\end{enumerate}

>>>>

Galaxy clusters have for many years been proposed as sources of gamma rays. If
shock acceleration in the ICM is an efficient process, a population of highly
relativistic cosmic-ray (CR) hadrons, in the form of protons and heavy ions, in
the ICM is to be expected. The main energy-loss mechanism for CR hadrons at high
energies is pion production through the interaction of CRs with nuclei in the
ICM. Pions are short lived and decay; the charged pions to electrons and
positrons,\footnote{A charged pion decays to a muon or antimuon, which
  subsequently decay to an electron or positron as well as neutrinos.} and the
neutral pion to gamma rays. Due to the low density of the ICM
($n_{\mathrm{ICM}}\sim 10^{-3}$ cm$^{-3}$), the large scales of a cluster and
the volume filling magnetic fields in the ICM, CR hadrons will be confined in
the cluster on timescales comparable and longer than the Hubble time
\citep[][]{article:Volk_etal:1996, article:Berezinsky_etal:1997} allowing them
to accumulate over time.  Depending on the CR pressure, quantified by the
CR-to-thermal pressure fraction $X_\CR=\expval{
  P_{\CR}}/\expval{P_{\mathrm{th}}}$, an observable flux of high-energy gamma
rays can be expected.

@@

A third mechanism for gamma-ray production in a galaxy cluster is
self-annihilation of a DM particle, e.g. a weakly interacting massive particle
(WIMP). As already mentioned, about 80\% of the cluster mass is in the form of
dark matter. Recent works \citep{article:EvansFerrerSarkar:2004,
  article:BergstromHooper:2006} have shown that the gamma-ray emissivity from DM
annihilations is expected to be proportional to the square of the DM density
integrated along the line of sight, making clusters of galaxies potentially good
candidates for DM searches \citep{article:PinzkePfrommerBergstrom2009,
  2011ApJ...726L...6C}. This despite their large distances compared to other
common targets for DM searches, such as dwarf spheroidal galaxies
\citep{article:Strigari_etal:2007, article:Acciari_etal:2010,
  article:Aliu_etal:2009} or the Galactic center
\citep{article:Kosack_etal:2004, article:Aharonian_etal:2006,
  article:Aharonian_etal:2009b}.

@@

The Coma cluster of galaxies (ACO 1656) is one of the most thoroughly studied
clusters over all wavelengths \citep{article:Voges_etal:1999}. Located at a
distance of about 100 Mpc \citep[$z=0.023$;][]{article:StrubleRood:1999}, it is
one of the closest massive clusters \citep[$M \sim
10^{15}M_{\odot}$;][]{article:Smith:1983, article:Kubo_etal:2008}. It hosts both
a giant radio halo
\citep{article:Giovannini_etal:1993,article:Thierbach_etal:2003} and peripheral
radio relic, which appears connected to the radio halo with a ``diffuse'' bridge
\citep[see discussion in][]{article:BrownRudnick:2010}. It has been suggested
that the relic is an infall shock \citep{article:Ensslin_etal:1998} and
successively demonstrated by cosmological simulations that model the non-thermal
emission processes \citep{article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008, article:Battaglia_etal:2009,
  article:Skillman_etal:2011}. Extended soft thermal X-ray (SXR) emission is
evident from the ROSAT all-sky survey in the 0.1 to 2.4 keV band
\citep{article:BrielHenryBohringer:1992} and observations with XMM-Newton
\citep{article:Briel_etal:2001} revealed substructure in the X-ray halo with
substantial turbulent pressure support that shows a lower limit of $\sim 10 \%$
of the total intracluster medium pressure \citep{2004A&A...426..387S}. As such it
is a natural candidate for gamma-ray observations.

In this article, results from the VERITAS observations of the Coma cluster of galaxies are reported, 
together with the resulting constraints on cosmic-ray particle populations, magnetic fields, and dark matter in the cluster. Throughout the analysis, a present day Hubble constant of $H_{0} = 100h$ km s$^{-1}$ Mpc$^{-1}$ with $h=0.7$ has been used.

@@

\item In the last step, we solve for $B_{0}$ subject to the boundary condition to match the observed synchrotron profile while fixing the profile of $X_{\CR}(r)$ as determined through the previous two steps. Note that for $B_{0} \gg B_{\rmn{CMB}}$ and a radio spectral index of $\alpha_{\nu}=1$, the solution would be degenerate.
\end{enumerate}
